Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2022-10-06 paper-correlations.tex
Map approach (1/2)

\section{The map approach.}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform operator and $N(r)$ is a total number of trials (vectors
thrown into sample).

In this approach we actively use fast Fourier transform (FFT)
to compute some correlation function ($S_2, F_{ss}, F_{sv}, C_2$)
in all directions.
The FFT algorithm let us compute auto and cross-correlation fast,
especially on GPU.

Lineal-path function doesn't have FFT realization,
therefore, we use different approach.

@@

Align and rotate --- simple operations with linear computational complexity.
Therefore, the complexity of the entire algorithm depends linearly 
on the number of directions and on the size of the image.

\subsection{Correlation map}

\subsubsection{Auto- and cross-correlation using FFT}

All subsequent functions are auto- or cross-correlations.
Therefore, here we describe a method to compute them using FFT.

\begin{algorithmic}[1]
  \Procedure{cross\_correlation}{$a, b$}
    \State $A \gets rfft(a)$\Comment{Real fast Fourie transform}
    \State $B \gets rfft(b)$
    \ForAll{$i \in CartesianIndices(A)$}
      \State $C[i] \gets conj(A[i]) \cdot B[i]$
    \EndFor
    \State $c = irfft(C)$ 
    \Comment{inverse rfft}
    \State \textbf{return} $c$
  \EndProcedure
\end{algorithmic}

One can use cross-correlation to compute auto-correlation,
but special function would permorm better.

\begin{algorithmic}[1]
  \Procedure{auto\_correlation}{$a$}
    \State $A \gets rfft(a)$
    \ForAll{$i \in CartesianIndices(A)$}
      \State $C[i] \gets |A[i]|^2$
    \EndFor
    \State $c = irfft(C)$ 
    \State \textbf{return} $c$
  \EndProcedure
\end{algorithmic}

CartesianIndices --- all indices of multidimensional array,
therefore, the for loop describe elementwise operations
and have linear computational complexity $O(n)$.
For both functions computational complexity is $O(n\ log(n))$.

It is worth noting, that the FFT approach imply
periodic boundary conditions by design.

\subsubsection{Total number of trials}

$auto\_correlation$ and $cross\_correlation$ return non-negative number
of success trials \cref{eq:probability}.
To compute probability we need to count total number of trials.
This value depends on boundary conditions.
We realize two types: periodic and non-periodic (flat).

@@

\subsubsection{Two-point funciton}

We realize two boundary conditions: periodic and non-periodic (flat).
For non-periodic boundary conditions one need to double image size
by padding new values with zeros.
Thus, if original image is $(m \times n)$,
then expanded image will have size $(2m - 1 \times 2n - 1)$.

\begin{algorithmic}[1]
  \Procedure{s2\_map}{$a, phase, periodic$}
    \State $a_1 \gets a = phase$
    \If{periodic}
      \State $a_2 \gets a_1$
    \Else
      \State $a_2 \gets expand(a_1)$\Comment{double image size by padding new values with zeros}
    \EndIf

    \State $success \gets AUTO\_CORRELATION(a_2)$

    \State $total \gets TOTAL(a, periodic)$

    \ForAll{$i \in CartesianIndices(success)$}
      \State $s2[i] \gets success[i]\ /\ total[i]$
    \EndFor

    \State \textbf{return} $s2$
  \EndProcedure
\end{algorithmic}

if n is the number of pixels, then computational complexity
of this funciton is $O(n\ log(n))$.

\subsubsection{Phase cross-correlation funciton}

\begin{algorithmic}[1]
  \Procedure{cc\_map}{$a, b, periodic$}
    \If{periodic}
      \State $a_1 \gets a$
      \State $b_1 \gets b$
    \Else
      \State $a_1 \gets expand(a)$
      \State $b_1 \gets expand(b)$
    \EndIf

    \State $success \gets CROSS\_CORRELATION(a_1, b_1)$

    \State $total \gets TOTAL(a, periodic)$

    \ForAll{$i \in CartesianIndices(success)$}
      \State $cc[i] \gets success[i]\ /\ total[i]$
    \EndFor

    \State \textbf{return} $cc$
  \EndProcedure
\end{algorithmic}

if n is the number of pixels, then computational complexity
of this funciton is $O(n\ log(n))$.

\subsection{Surface-surface function}

Computation of surface-surface function consists of two steps 
\textcolor{red}{[link to surface paper]}:
\begin{enumerate}
  \item extracting surface $M = \nabla I(a)$
  \item computing $F_{ss} = S_2(M)$
\end{enumerate}

For surface extraction we use `imgradient' function
and Sobel kernel as default kernel, 
both defined in ImageFiltering.jl package \textcolor{red}{[add link]}.

\begin{algorithmic}[1]
  \Procedure{surf\_surf}{$a, phase, periodic, kernel$}
    \State $a_1 \gets a = phase$
    \State $M \gets imgradient(a_1, kernel)$
    \State $Fss = S2\_MAP(M, periodic)$
    \State \textbf{return} $Fss$
  \EndProcedure
\end{algorithmic}

if n is the number of pixels, then computational complexity
of this funciton is $O(n\ log(n))$.

\subsection{Surface-void function}
Computation of surface-void function consists of three steps 
\textcolor{red}{[link to surface paper]}:
\begin{enumerate}
  \item extracting surface $M = \nabla I(a)$
  \item extracting void phase $V = I_v(a)$
  \item computing $F_{sv} = cross\_correlation(M, V)$
\end{enumerate}

\begin{algorithmic}[1]
  \Procedure{surf\_void}{$a, phase, periodic, kernel$}
    \State $a_1 \gets a = phase$
    \State $M \gets imgradient(a_1, kernel)$
    \State $V \gets a = 0$
    \State $Fsv = CC\_MAP(M, V, periodic)$
    \State \textbf{return} $Fsv$
  \EndProcedure
\end{algorithmic}

if n is the number of pixels, then computational complexity
of this funciton is $O(n\ log(n))$.

\subsection{Cluster funciton}

Computation of cluster function consists of three steps
\begin{enumerate}
  \item label different clusters with distinct integers
  \item compute $S_2$ map for every cluster
  \item add up all maps
\end{enumerate}

For labeling we use `label\_components' function from
Images.jl package \textcolor{red}{[add link]}.

\begin{algorithmic}[1]
  \Procedure{cluster\_map}{$a, phase, periodic$}
    \State $a_1 \gets a = phase$
    \State $labeled\_a \gets label\_components(a_1, periodic)$
    \State $number\_of\_labels = maximum(labeled\_a)$
    \ForAll{$label \in range(1, number\_of\_labels)$}
      \State $b \gets labeled\_a = label$
      \If{label = 1}
        \State $c2 \gets S2\_MAP(b, periodic)$
      \Else
        \State $c2 \gets c2 + S2\_MAP(b, periodic)$
      \EndIf
    \EndFor
    \State \textbf{return} $c2$
  \EndProcedure
\end{algorithmic}

if n is the number of pixels and N is the number of clusters, then computational complexity
of this funciton is $O(N\ n\ log(n))$.

>>>>

\section{The map approach.}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into sample). If we want to calculate $S_2$ function in periodic mode we
need only to apply \cref{eq:s2-ft} to an input to obtain the result because DFT
implicitly continues a signal to infinity \textcolor{red}{Does this need a
clarification?}. In non-periodic mode we need to pad an input with zeros to at
least twice the size minus one for each dimension. In periodic mode a total
number of trials is independent of $r$ and equals to the number of elements in
an input array, or otherwise is calculated as described
\textcolor{red}{Here}. The following algorithm summarizes all said above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}| / N$
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation if $O(M \log M)$ where $M$ is the
number of elements in the input.

This function is a special case of a cross-correlation function $\rho_{ij}$
which computes a correlation between phases $i$ and $j$ using a formula
\begin{equation}
  \rho_{ij}(r) = \frac{F^{-1} [F[A^{(i)}](z) \overline{F[A^{(j)}](z)}] (r)}{N(r)} \label{eq:cross-ft}
\end{equation}
or an equivalent algorithm
\begin{algorithmic}[1]
  \Procedure{$\rho_{ij}$}{$A, i, j, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(i)} \gets I^{(i)} (A)$
  \State $A^{(j)} \gets I^{(j)} (A)$
  \State $\hat{A}^{(i)} \gets F[A^{(i)}]$
  \State $\hat{A}^{(j)} \gets F[A^{(j)}]$
  \State $\hat{\rho} \gets \hat{A}^{(i)} \overline{\hat{A}^{(j)}} / N$
  \State \textbf{return} $F^{-1} [\hat{\rho}]$
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated in the similar manner. Firstly, we apply an
indicator function to an input $A$: $A^{(i)} \xleftarrow{I^{(i)}} A$. When we
label connected components in $A^{(i)}$ using one of many popular algorithms for
this purpose \textcolor{red}{cite something}. This algorithm must take into
account if the function is calculated in periodic mode or not. Then $C_2$
function is a sum of $S_2$ functions calculated for all clusters
separately. Algorithmic complexity of this implementation is $O(MC \log M)$
where $M$ is the number of elements in the input and $C$ is the number of
clusters. The algorithm is as follows:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

For surface-surface and surface-void functions we use an edge detection filter
and then calculate either autocorrelation of the edge or cross-correlation of
the edge and the void phase. The edge is extracted by convolving an input with a
short signal $E$ (\cref{eq:filter-1d} for 1D case, \cref{eq:filter-2d} for 2D case or
\cref{eq:filter-3d-1}-\cref{eq:filter-3d-3} for 3D case).
\begin{align}
  E^{1D} &= \frac{\sqrt{2}}{3} \left[
    \begin{array}{ccc}
        1 & -2 & 1
      \end{array}
    \right] \label{eq:filter-1d} \\
  E^{2D} &= \frac{\sqrt{2}}{9} \left[
    \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -8 & 1 \\
        1 & 1 & 1
      \end{array}
    \right] \label{eq:filter-2d} \\
  E^{3D}_{ij0} &= \frac{\sqrt{2}}{27} \left[
    \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3d-1} \\
  E^{3D}_{ij1} &= \frac{\sqrt{2}}{27} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -26 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3d-2} \\
  E^{3D}_{ij2} &= F^{3D}_{ij0} \label{eq:filter-3d-3}
\end{align}

An algorithm for $F_{ss}$ function is the following:
\begin{algorithmic}[1]
  \Procedure{$F_{ss}$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $E_A \gets E * A^{(phase)}$
  \State $\hat{E}_A \gets F[E_A]$
  \State $\hat{F}_{ss} \gets |\hat{E}_A| / N$
  \State \textbf{return} $F^{-1} [\hat{F}_{ss}]$
  \EndProcedure
\end{algorithmic}

@@

Align and rotate are simple operations with linear computational complexity.
Therefore, the complexity of the entire algorithm depends linearly 
on the number of directions and on the size of the image.

\subsubsection{Total number of trials}
$auto\_correlation$ and $cross\_correlation$ return non-negative number
of success trials \cref{eq:probability}.
To compute probability we need to count total number of trials.
This value depends on boundary conditions.
We realize two types: periodic and non-periodic (flat).

@@
